Abstractions for Multi-Sorted Substitutions

### 2023-09-20 paper/paper.tex
Update.

  In the context of multi-sorted syntax, the notion of a type can be
  described as a mapping between sorts.
  For System F, the expression sort \ACode{\ACon{𝕖}} maps to the type sort
  \ACode{\ACon{𝕥}} and the sort \ACode{\ACon{𝕥}} maps to the kind sort
  \ACode{\ACon{𝕜}}.
  To teach our framework about types, we have to instantiate the following structure:
  \LibCode\KTypeSorts
  For System F the instantiation is
  \AppCode\FTypes
  For a nicer interface, we define a type analogously to a term:
  \LibCode\KTypes

  \subsection{Typing Context}
  A typing context is then a function which maps variables to types.
  As we want our framework to support dependent types, the types returned by
  a typing context are allowed to use the domain of the 
  To be able to deal with dependent types, the result types are only allowed
  to refer to the variables that come before 
  \LibCode\KContextHelper
  \LibCode\KContexts
  \LibCode\KContextLookup

  \LibCode\KVariableTyping

  \section{Typing}
  \label{sec:typing}
  Equipped with a notion of types and type context we are ready to define
  the multi-sorted typing relation for System F, incorporating both
  typing and kinding:
  \AppCode\FTyping
  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing of the lambda abstraction \ACode{\ACon{⊢λ}} weakens the
    type of the codomain \ACode{t₂}.
    This is necessary, because with multi-sorted syntax, types could
    also depend on expressions, so the sub typing derivation accounts for
    another variable which is not used by the type.
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-scoped due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  and instantiate it for System F:
  \AppCode\FTypingInst

  Next we define \ACode{\ACon{TypingKit}}s, which are analog to
  \ACode{\ACon{Kit}}s, but abstract about variable and term
  \emph{typing}, instead of variables and terms, themselves.
  \LibCode\KTypingKit
  Building on the fields of a \ACode{\ACon{TypingKit}}, we define further operations:
  \begin{itemize}
  \item
    Using the variable/term typing, we can define a renaming/substitution typing:
    \LibCode\KMapTyping
  \item 
    Lifting a map preserves its typing:
    \LibCode\KLiftTyping
    % \LibCode\KLiftTypingProof
  \item 
    If a variable/term has a typing, then so does its singleton renaming/substitution:
    \LibCode\KSingleTyping
    % \LibCode\KSingleTypingProof
  \end{itemize}
  To make it easier to explicitly pass typing kits,
  we define the notation \ACode{Γ₂ \ADef{∋*/⊢*[} TK \ADef{]} ϕ \ADef{∶} Γ₁}
  to stand for \ACode{Γ₂ \ADef{∋*/⊢*} ϕ \ADef{∶} Γ₁} given some
  \ACode{\ACon{TypingKit}} \ACode{TK}.
  % \LibCode\KTypingNotation

  We then define a \ACode{\ACon{TypingTraversal}} analogously to
  \ACode{\ACon{Traversal}}, but instead of defining the application of
  maps, it defines the type preservation of applying maps:
  \LibCode\KTypingTraversal
  Given a term \ACode{e} with typing \ACode{⊢e} and
  a renaming/substitution \ACode{ϕ} with typing \ACode{⊢ϕ},
  the term \ACode{⊢e \AField{⊢⋯} ⊢ϕ} is a typing for \ACode{e \AField{⋯} ϕ}.

@@

  \AppCode\FPreserve
  The interesting parts of the proof are:
  \begin{itemize}
  \item 
  \end{itemize}

  \AppCode\FTypingTraversal

@@

  We define a standard small-step operational semantics for System F
  with full reduction:
  \AppCode\FReduction

  Proving subject reduction is now straightforward:
  \AppCode\FSubjectReduction
  \AppCode\FSubjectReductionProofInteresting
  We only listed the cases that do not follow directly from the
  induction hypothesis:
  \begin{itemize}
  \item The beta-reduction rule for type-application, \ACode{\ACon{β-Λ}},
    is a direct consequence of substitution preserves typing.
    \ACode{⊢e₂} is the typing for \ACode{e₂}.
    \ACode{⊢⦅ ⊢e₂ ⦆} is the substitution-typing for the singleton substitution 
    \ACode{\AField{⦅ e₂ ⦆}}.
    \ACode{\AField{⊢e₁ ⊢⋯ ⊢⦅ ⊢e₂ ⦆}} states that applying the singleton substitution
    to \ACode{e₁} preserves typing.
  \item The beta-reduction rule for value-application is almost the same,
    but requires us to addionaly apply an equality to the type index.
  \end{itemize}

@@

  \section{Reflection \& Generics}
  \label{sec:reflection}
  We used Agda's reflection mechanism\cite{TODO} to derive
  instantiations related to all structures for untyped substitition,
  i.e. \ACode{\ACon{Syntax}}, \ACode{\ACon{Traversal}} and
  \ACode{\ACon{ComposeTraversal}}.

  To gain insight into the class of object languages supported by our
  reflection algorithm, we have formalized a class of object languages
  using a generic programming technique and instantiated the structures
  generically. Informally, this class contains all object languages which
  \begin{enumerate}
  \item have a variable constructor of type TODO;
  \item use subterms only directly (e.g. not in lists);
  \item only extend the scope-context of subterms.
  \end{enumerate}

@@

  We were able to describe many languages in this shape, including
  session types and type state.

  We believe restriction 2 is purely technically and can be lifted
  with a more sophisticated reflection algorithm, as parameterized
  datastructures can also be inlined into the syntax definition as terms
  of a new sort.

@@

  \section{Case Studies}
  \label{sec:casestudies}
  \begin{itemize}
  \item
    For lambda calculus with dependent function types, the framework
    just works out of the box for both deriving and instantiating the
    \ACode{\ACon{TypingTraversal}}. Additionally, having the
    \ACode{\ACon{ComposeKit}} abstraction turned out to be useful
    to prove confluence, allowing to unify certain lemmas, which
    otherwise need to be proved for both renamings and substitutions,
    e.g. that for two terms \ACode{t} and \ACode{t'}
    with \ACode{t \ACon{↪} t'} it follows that \ACode{(t \AField{⋯} ϕ)
    \ACon{↪} (t' \AField{⋯} ϕ)} for any map \ACode{ϕ}.
  \item
    For System F with subtyping the main challenges are how to represent
    subtyping constraints and how to deal with the fact that
    substitution preserves typing is not generally true, as type variables
    have subtyping bounds that need to be respected.
    Instead of defining typing contexts and type preservation lemmas independent
    of our framework, we found a way that still allows us to benefit
    from the typing machinery.
    When binding a type variable with a subtyping constraint \ACode{α <: t},
    we first bind the type variable as \ACode{α : ★}, and then bind the constraint
    with a dummy name \ACode{\_ ∶ (α <: t)}. This description is
    similar to first-class constraints, but restricted enough to be
    isomorphic to the original formalization.
    With that encoding substitution preserves typing is generally true again, enabling
    the use of the framework's typing contexts and \ACode{\ACon{TypingKit}} machinery.
    Replacing a type variable \ACode{α <: t} with a type \ACode{t'}
    which is not a subtype of \ACode{t'}, the resulting term is still
    well-typed, but in a context with an insatisfiable constraint.
  \item
    Object language with pattern matching pose an interesting problem.
    A clause \ACode{p ⇒ e} in a match-expression consists of a pattern
    \ACode{p} and a continuation expression \ACode{e}.
    The pattern \ACode{p} can bind arbitrary many variables in subpatterns,
    which are then available as free variables in \ACode{e}.
    % The challenge is to track the sorts of variables bound by a pattern,
    % to make them available to \ACode{e} by extending \ACode{e}'s sort
    % context correspondingly.
    To track the sorts of variables bound by a pattern, one can define
    the sort of patterns to contain the sorts of the variables that it binds:
    % 𝕡 : List (Sort Var) → Sort NoVar
    The syntax for variable patterns and clauses can then be specified as:
    % varpat : S ⊢ 𝕡 [ 𝕖 ]  -- pattern binding an expression variable
    % pairpat : S ⊢ 𝕡 S₁ → S ⊢ 𝕡 S₂ → S ⊢ 𝕡 (S₁ ++ S₂)
    % clause : S ⊢ 𝕡 S' → (S' ++ S) ⊢ 𝕖 → S ⊢ 𝕔
  \end{itemize}

>>>>

  In the context of multi-sorted syntax, the notion of a type can be
  described as a mapping between sorts.
  For System F, the expression sort \ACode{\ACon{𝕖}} maps to the type sort
  \ACode{\ACon{𝕥}} and the sort \ACode{\ACon{𝕥}} maps to the kind sort
  \ACode{\ACon{𝕜}}.
  The following structure is used to teach our framework about types:
  \LibCode\KTypeSorts
  For System F the instantiation is
  \AppCode\FTypes
  There are two things to discuss:
  \begin{enumerate}
  \item
    The \ACode{\AField{↑ᵗ}} function maps a sort of arbitrary sort type,
    to a sort of a potentially different sort type, which is expressed
    by the use of an existential.
    For System F we require this generality, as the sort
    \ACode{\ACon{𝕥}} can have variables, whereas its corresponding type
    sort \ACode{\ACon{𝕜}} cannot.
  \item
    Some sorts, like \ACode{\ACon{𝕜}}, do not have corresponding type
    sorts, but we still need to provide one, as \ACode{\AField{↑ᵗ}} is a
    total function.
    For such sorts, we can simply use arbitrary sort types, as the
    formalization will never use them.
  \end{enumerate}
  To hide the existential, we define \ACode{S \ACon{∶⊢} s}, which represents
  the type for a term \ACode{S \ACon{⊢} s}.
  \LibCode\KTypes

  \subsection{Typing Context}
  Equipped with a notion of types, we define a typing context as a
  function that maps variables to types of the same sort.

  As we want our framework to support dependent types, we allow a type returned
  by the typing context to use all variables bound previously in the typing context.
  This is achieved by the \ACode{drop-∈} function, which cuts of all
  sorts from \ACode{S} that are at smaller or equal positions than a variable \ACode{x}.

  % \LibCode\KContextHelper
  \LibCode\KContexts
  To use a type returned from a typing context, it needs to be
  weakened appropriately through the \ACode{\ADef{wk-telescope}} function:
  \LibCode\KContextLookup
  We then define a variable typing as followed:
  \LibCode\KVariableTyping

  \section{Typing}
  \label{sec:typing}
  Equipped with a notion of types and type context, we are ready to define
  the multi-sorted typing relation for System F, incorporating both
  typing and kinding:
  \AppCode\FTyping
  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor, which covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing \ACode{\ACon{⊢λ}} of the lambda abstraction, which
    weakens the type of the codomain \ACode{t₂}.
    This is necessary, because multi-sorted syntax allows types
    to depend on expressions, so the typing derivation for \ACode{e}
    has to account for a variable, which is not used by the type.
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-scoped due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  The instantiation for System F is straightforward:
  \AppCode\FTypingInst

  \subsection{An Abstraction for Type Preservation}
  By now, the reader probably knows what comes next: we build an
  abstraction to unify type preservation for renamings and substitutions,
  eliminating the dependencies by yet another type of kits.

  We start with \ACode{\ACon{TypingKit}}s, which abstract over
  variable and term \emph{typing}, and then define a
  \ACode{\ACon{TypingTraversal}}, which provides substitution
  preserves typing for all \ACode{\ACon{TypingKit}}s.
  \LibCode\KTypingKit
  The operations are completely analog to those of a \ACode{\ACon{Kit}}, but talk
  about typing instead of terms.
  Building on the fields of a \ACode{\ACon{TypingKit}}, we define further operations:
  \begin{itemize}
  \item
    Using the variable/term typing, we can define a renaming/substitution typing:
    \LibCode\KMapTyping
    \ACode{Γ₂ \ADef{∋*/⊢*} ϕ \ADef{∶} Γ₁} states that \ACode{ϕ} is a map that takes
    terms in \ACode{Γ₁} to terms in \ACode{Γ₂}.
  \item 
    Lifting a map preserves its typing:
    \LibCode\KLiftTyping
    % \LibCode\KLiftTypingProof
  \item 
    If a variable/term has a typing, then so does its singleton renaming/substitution:
    \LibCode\KSingleTyping
    % \LibCode\KSingleTypingProof
  \end{itemize}
  To make it easier to explicitly pass typing kits,
  we define the notation \ACode{Γ₂ \ADef{∋*/⊢*[} TK \ADef{]} ϕ \ADef{∶} Γ₁}
  to stand for \ACode{Γ₂ \ADef{∋*/⊢*} ϕ \ADef{∶} Γ₁} given some
  \ACode{\ACon{TypingKit}} \ACode{TK}.
  % \LibCode\KTypingNotation

  We then define a \ACode{\ACon{TypingTraversal}} analogously to
  \ACode{\ACon{Traversal}}, but instead of defining the application of
  maps, it defines the type preservation of applying maps:
  \LibCode\KTypingTraversal
  Given a term \ACode{e} with typing \ACode{⊢e} and
  a renaming/substitution \ACode{ϕ} with typing \ACode{⊢ϕ},
  the term \ACode{⊢e \AField{⊢⋯} ⊢ϕ} is a typing for \ACode{e \AField{⋯} ϕ}.
  The large amount of kit-parameters does not impose any restriction,
  as both our \ACode{\ACon{Kit}}s support the \ACode{\ACon{WkKit}}
  extension and can be composed arbitrarily. They only make the previously
  built machinery available to the proof of \ACode{\AField{\_⊢⋯\_}}.

@@

  \AppCode\FPreserve
  The interesting parts of the proof are:
  \begin{itemize}
  \item
    There is a strong similarity to the instantiation of
    map traversal \ACode{\AField{\_⋯\_}}. Where \ACode{\AField{\_⋯\_}}
    used \ACode{\AField{`/id}} or \ACode{\AField{\_↑\_}}, we now use their
    preservation lemmas \ACode{\AField{⊢`/id}} or \ACode{\AField{\_∋↑/⊢↑\_}}.
  \item
    The lambda typing constructor \ACode{\ACon{⊢λ}} weakens the type \ACode{t₂} to
    shield it from expression-substitution, requiring us to use the
    \ACode{\AField{⋯-↑-wk}} lemma.
  \item
    The type application constructor \ACode{\ACon{⊢∙}} substitutes \ACode{t₁} into \ACode{t₂},
    requiring us to use the \ACode{\AField{dist-↑-⦅⦆}} lemma.
  \end{itemize}
  Finally, we instantiate the \ACode{\ACon{TypingTraversal}} structure:
  \AppCode\FTypingTraversal

@@

  To finish our running example, we define a standard small-step
  operational semantics for System F with full reduction and prove subject reduction:
  \AppCode\FReduction
  \vspace{0mm}
  \AppCode\FSubjectReduction
  \AppCode\FSubjectReductionProofInteresting
  We only listed the cases for $\beta$-reduction, as the other cases
  follow directly from the induction hypothesis:
  \begin{itemize}
  \item The beta-reduction rule for type-application, \ACode{\ACon{β-Λ}},
    is a direct consequence of substitution preserves typing.
  \item The beta-reduction rule for value-application is almost the same,
    but requires us to additionally show that the type is not affected by the
    expression substitution. This is a direct consequence of the weakening
    used in \ACode{\ACon{⊢λ}}.
  \end{itemize}

@@

  \section{Reflection \& Generics}
  \label{sec:reflection}
  We used Agda's reflection mechanism to derive
  instantiations related to all structures for untyped substitition,
  i.e. \ACode{\ACon{Syntax}}, \ACode{\ACon{Traversal}} and
  \ACode{\ACon{ComposeTraversal}}.

  To gain insight into the class of object languages supported by our
  reflection algorithm, we have formalized a class of object languages
  using a generic programming technique and instantiated the structures
  generically. Informally, this class contains all object languages described
  via multi-sorted syntax, which
  \begin{enumerate}
  \item have a variable constructor of type
    \ACode{S \ACon{∋} s \AKw{→} S \ACon{⊢} s};
  \item use subterms only directly (e.g. not in lists);
  \item only extend the scope-context of subterms, but never modify it.
  \end{enumerate}

@@

  We believe restriction 2 is purely technical and can be lifted
  with a more sophisticated reflection algorithm, as parameterized
  datastructures can also be inlined into the syntax definition as terms
  of a new sort.

@@

  \section{Case Studies}
  \label{sec:casestudies}
  Using our full implementation of the framework, we proved subject
  reduction for the following object languages:
  \begin{itemize}
  \item
    For lambda calculus with dependent function types, the framework
    just works out of the box for both deriving and instantiating the
    \ACode{\ACon{TypingTraversal}}.
    Furthermore, the \ACode{\ACon{ComposeKit}} abstraction turned out
    to be useful to prove confluence, allowing to unify certain
    lemmas, which otherwise need to be proved for both renamings and
    substitutions.
    % e.g. that for two terms \ACode{t} and \ACode{t'}
    % with \ACode{t \ACon{↪} t'} it follows that \ACode{(t \AField{⋯} ϕ)
    % \ACon{↪} (t' \AField{⋯} ϕ)} for any map \ACode{ϕ}.
  \item
    For System F with subtyping the main challenges are how to represent
    subtyping constraints and how to deal with the fact that
    substitution preserves typing is not generally true, as type variables
    have subtyping bounds that need to be respected.
    While it would be possible to use our framework only to derive
    untyped substitution and define typing contexts and type preservation
    lemmas by hand, we found an encoding, which still allows us to use the
    the \ACode{\ACon{TypingKit}} abstraction.
    Instead of binding a type variable with a subtyping constraint
    \ACode{α <: t}, we first bind the type variable as \ACode{α : ★}, and
    then bind the constraint as \ACode{c ∶ (α <: t)}. This description is
    similar to first-class constraints, but restricted enough to be
    isomorphic to the original formalization, as constraint variables
    cannot be accesses by the user.
    With that encoding substitution preserves typing is generally true
    again, enabling the use of the framework's typing contexts and
    \ACode{\ACon{TypingKit}} machinery.
    When replacing a type variable \ACode{α <: t} with a type \ACode{t'},
    which is not a subtype of \ACode{t'}, then the resulting term is still
    well-typed, but in a context with an unsatisfiable constraint.
  \item
    Object languages with pattern matching can be encoded by adding
    the sorts of the variables bound by a pattern to the pattern sort
    \ACode{\ACon{𝕡}} itself:
    \begin{quote}
    \ACode{\ACon{𝕡} \AKw{:} \ACon{List (Sort Var)} \AKw{→} \ACon{Sort NoVar}} 
    \end{quote}
    This allows the clause of a \ACode{\ACon{match}} expression to
    read the bound variables from the pattern sort and make them
    available as free variables in its body expression.
  \end{itemize}
